The impact of earthquake cycle variability on neotectonic and paleoseismic slip rate estimates

### 2018-11-26 manuscript/reviews/response_letter.md

---
title: |
  Response to Reviewers, "The impact of earthquake cycle variability on 
  neotectonic and paleoseismic slip rate estimates"
author: 
- name: Richard Styron
---

@@

## Review by Michael Oskin

> [O1] This paper presents a useful thought experiment on the impact of 
> earthquake cycle variability on measured slip rates, and concludes that the 
> convergence on the expected value is a function of the coefficient of 
> variation. Overall this is a sensible conclusion. Underpinning this analysis 
> are four assumed variants of earthquake recurrence, and a function to express 
> the variability of slip per event.

> [O2] I would like to see the effect of COV isolated from the slip per event 
> distribution (use 1m slip for every event).

I performed this experiment; the figure with the slip rate results is shown
below for comparison with Figure 5 in the paper, and are included in a new
document in the supplemental material. The most relevant figure is also shown
here (Figure \ref{nsv_sre})

![Slip rates with no displacement variability \label{nsv_sre}](../figures/nsv_slip_rate_envelopes.pdf)

The differences between the results of this experiment (fixed per-event
displacement of 1 m) and of the numerical experiment performed in the manuscript
are basically that these results are less smooth, but the total variance at any
point in time (*x*-axis) is less. This is simply due to removing the
stochasticity from one of the two variables in the system. The relative
spread in the data and convergence rates are unchanged. This is to be expected
as the variability in the per-event displacement is the same for all recurrence
distributions, so even though it is a random variable in the simulations, it is
not an experimental variable.

@@

[Note: I've analyzed Weldon's data (from K. Scharer's refined OxCal earthquake 
dates) and found that earthquake recurrence interval duration at the Wrightwood 
and Pallett Creek sites has a negative autocorrelation, i.e. a short recurrence 
interval is likely to be followed by a long recurrence interval, and vice 
versa. This is quite unlike the autocorrelation in the Puget Lowlands of
Washington State, which is a network of generally similar faults; here 
autocorrelation is positive, so a long recurrence interval will more likely be 
followed by another long recurrence interval. I don't know what this means, or 
whether it's all noise, but it's intriguing to me and I put this bit in the 
paper in hopes of catching the attention of others who may be looking for a 
project.]

*Changes: None.*

> [O20] Page 6, line 1. Zero friction at rupture arrest is very unrealistic, 
> and not a prerequisite for characteristic behavior.

It is not a necessary condition but it is a sufficient one, when coupled with
fairly regular reloading and failure conditions. I personally think it's
unlikely as well (see Styron and Hetland 2015 for example), but zero (or very
near zero) friction is part and parcel with complete stress drop in major
earthquakes, which is supported by many studies (e.g., Hardebeck, 2012;
Hasegawa et al., 2011) though I am pretty suspicious of the results.
Furthermore, there are a host of laboratory experiments which suggest that
friction during slip decreases to very low values (e.g., Di Toro et al.,
2004).

@@

It was a LaTeX error on my part, and has been fixed.

*Changes: LaTeX fix.*

> [A5] P1L1:  The connection to locking depth should be explained a bit more.  
> Good to also provide a reference here.

I added a phrase stating that the width of the zone affected by
earthquake-cycle strains is a function of the fault's locking depth, and
added refs to Savage and Burford (1973), the classic
locked-fault-above-creeping-fault reference, and Hetland and Hager (2006),
which demonstrates that this process can instead be the result of
post-seismic relaxation.

@@

I have added another numerical experiment using an empirical slip distribution; 
see the response to [06a].

*Changes: numerical experiment w/ alternate slip distribution added.*

> [A8] P5L17:  Here you describe qualitatively how more or less closely the 
> different distributions align with the mean slip rate.  While doing this 
> qualitatively is ok to first order, I suggest that you go one step further 
> and compute some form of misfit function i.e., residual (simplest a L1 or L2 
> norm).

An L-norm of any sort isn't appropriate because the numbers here aren't 
pairwise (observation, model) data. Instead, there is a single 'true' value at 
any time $t$, which is invariably 1, and then hundreds to thousands of 
simulated values at each time $t$.

@@

Instead, because the true value is 1, all of the results shown in Figure 5 are 
quantified percentiles of misfit. While I could calculate the CV as suggested 
in comment [A3], this is pretty reductive and the numbers at any time $t$ can 
be retrieved from Figure 5.

*Changes: None.*

> [A9] P6L2: stress drop doesn’t need to be “complete” –just has to be “the 
> same” each time to get to the outcome you describe here. Maybe better 
> rephrase accordingly.

No, this isn't true. The outcome that I describe is a *correlation* between 
loading time and displacement, not an *invariance* of either loading time or 
displacement. Having 'the same' stress drop doesn't predict correlated loading 
time/recurrence intervals and displacement.

@@

Since this review seems to aim to shame rather than help, the number of changes
are few.

> [B2a] A couple of omissions in this paper are particularly striking.

> [B2b] First, how do we have a paper addressed to “variability” in fault
> slip rate, addressing particularly the problem of small samples, without
> mentioning the methods of estimation for censored samples? There is an
> extensive statistical literature to estimate parameters and uncertainties and
> numerous recent papers applying it in paleoseismic contexts. This literature
> provides real quantitative methods to deal with the open intervals, long or
> short, that affect the geologist’s estimate of fault slip rate and recurrence
> estimation. These are real equations, with real uncertainties. One would look
> in vain in this paper for anything of similar substance.

The paper doesn't go into depth on the topics of open intervals and censored
data because this paper does not use any samples of earthquake recurrence
intervals or per-event displacements in the rate calculations. Though they
are present behind the scenes, all of the numerical results in the paper are
slip rates derived as the total offset over the total time (equation 1)--no
information on how many earthquakes are present in that time window, or what
the duration of open intervals may be, is used in the calculations. This is
intended to simulate a neotectonic-type slip rate estimation (as written in page
2, lines 24-25 of the original manuscript), where an offset geological marker
layer is dated and the offset is measured. I have reinforced this by
restating in the methods that neither the number of events or the length of
the closed or open intervals are known.

The decision to avoid discussing the topic of including censored data in
parameter estimation is also warranted because the most robust method to
incorporate open intervals are in maximum-likehood estimation, which is a
complex topic to bring up, as are survival analysis and related methods. Since
these topics are complimentary, not central, to this paper I don't see the
benefit of adding a dense paragraph (much less any equations) for methods that
aren't relevant.

@@

I don't disagree that more substance is offered by methods developed through the
decades by many researchers than I can offer in one short paper. Sorry!

*Changes: brief reinforcement that this study uses neotectonic (age/offset) slip
rate measurements, not those from multiple events as in a paleoseismic context.*

@@

> [B2c] Second, pages of this paper could be replaced (and improved) by a
> presentation and discussion of the properties of the standard error. E.g.,
> given an estimate a sample- based estimate of the mean, how far might the
> population (or true) mean be from the estimate? S.E. is estimated by the
> sample standard deviation divided by the square root of the number of
> samples. So, of course, estimates from small samples from a fuzzy log normal
> converge more slowly than from a well-defined (quasi-periodic) lognormal.
> Instead of a small equation (SE=s/sqrt(n)), our paper back-calculates the
> result using 2 million years of samples, and presents the results like a new
> discovery . And again, with little by way of meaningful uncertainties (e.g.,
> p1, lines 11- 13, 14- 16).

The standard error does share the same basic decrease in dispersion with
increasing number of events, but it is symmetric even though the distributions
used here are not, and it does not account well for the systematic bias in short
time windows that are observed in these results. The misfit between the
symmetric and asymmetric approximations becomes particularly bad when estimating
the 'epistemic' uncertainty as in Figure 6, which is the uncertainty that one
should apply to a measurement to account for earthquake-cycle variability. I
have calculated the standard error and added to this graph show in Figure
\ref{se} below (not added to manuscript). The results do not support the use
of the standard error here. The upper error envelope under-approximates the
uncertainty at short time windows (due to the measurement bias), but then
becomes acceptable. However, the lower error envelope stays at or below the
5th percentile, though it should be the 1-standard deviation approximation.

![Standard error (S.E.) plotted on top of Figure 6 from the manuscript. \label{se}](./epist_unc.pdf)

The standard error is also not a great fit for neotectonic slip rate
estimations: They are typically derived by dividing net offset over net time
$n$ is not known, nor are the durations of the open intervals.
It would be different if this paper was at all concerned with the calculation of
recurrence intervals, per-event displacements, or other parameters of individual
earthquakes where a number of samples is used, but that is a different topic.

Finally, meaningful uncertainties are given in Table 1 for all distributions at
a wide range of times. It's not appropriate to reduce all of these to a single equation.

*Changes: Short discussion of standard error added to 'slip rate calculations' section.*

@@

> [B3] 1 L 9,10: We read that the most important parameter is the coefficient of
>  variation. First, this equation is the arithmetic coefficient of variation,
>  and not the CV for a lognormal distribution. The CV of a lognormal does not
>  depend on the mean.

The reviewer is really confused about the lognormal distribution (see [B6] and
[B7] where the reviewer continues to make mistakes and level false accusations
at me).

I was quite worried that I made a mistake after reading this, so I re-checked
my math, my code and re-read a range of references (e.g., *Statistical
Distributions* by Evans, Hastings and Peacock, 1993; *Confidence Intervals
for the Coefficent of Variation for the Normal and Lognormal Distributions*,
Koopmans, Owen, and Rosenblatt, *Biometrika*, v. 51, no. 1/2, 1964, as well
as a lot of internet pages). Nowhere do I find alternative definitions of the
$CV$ for a lognormal distribution, except in the Wikipedia page for
Coefficient of Variation where two conflicting definitions for the Geometric
Coefficient of Variation are given. Both of these involve estimating the $CV$
from samples rather than as an algebraic transformation of the equations for
a lognormal distribution. The first definition is mathematically the same as
the standard definition; the difference lies in whether the estimation
procedure is done on the raw data or log-transformed data. The second is
simply the square of the first.

Just to reiterate, the arithmetic coefficient of variation for the lognormal
distribution is the standard definition. It is also equivalent to the geometric
coefficient of variation for a theoretical longormal distribution (although
not necessarily for sample-based estimates).

@@

The reviewer's confusion may stem from the fact that there are two equations
for the coefficient of variation that get regularly passed around. However, as I
demonstrate in the response to [B7c], they are mathematically equivalent. The
simple definition, which I use in the paper, is $CV = \frac{\sigma}{\mu}$,
where $\sigma$ and $\mu$ are the standard deviation and mean of the
distribution. The other definition, $CV=\sqrt{e^{\phi^2}-1}$, is given in
terms of the standard deviation $\phi$ of the natural log of the lognormal
distribution.

OK, this is pretty funny: All of the confusion that the reviewer has about
the lognormal distribution appears to be because of a misreading of the
Wikipedia pages for the lognormal distribution and for the coefficient of
variation. I mentioned the geometric CV (non-)issue above. The quip that the
'CV of a lognormal does not depend on the mean' is... kind of true? As
demonstrated in the response to [B7c], the standard deviation of a lognormal
is a *function of the mean* in that the mean is a coefficient in the standard
devation equation if it's expressed in the typical shape and scale
parameters. Therefore $CV = \sigma / \mu = \mu A /\mu$ where $A$ is another
mathematical expression (see [B7c]). So, yes, technically it is true but only
because $\mu$ is on the top and the bottom of the fraction and therefore cancels
out; furthermore, you can't derive the expression $A$ without knowing both
$\mu$ and $\sigma$. Regardless, the quip gives all outward appearance of
being a slight modification of the sentence "Contrary to the arithmetic
standard deviation, the arithmetic coefficient of variation is independent of
the arithmetic mean" which is below the equation for the $CV$ of a lognormal
on the Wikipedia page. It is not a statement that makes any practical sense
unless one is beginning with log-transformed data or distributions.

So to conclude, my math, code and analysis are correct in terms of the use of
the $CV$.

@@

*Changes: None.*

> [B4] We could stop here, but a central flaw in the paper is exposed – nothing
> in this paper addresses how to obtain this most important parameter. If
> attempted, the essential emptiness of a 2,000,000 year sample would emerge. No
> real data set in paleoseismology resolves the mean and standard deviation to
> better than maybe 50%. Typical sites do well to resolve it to a factor of 2.
> P.3, line 22- 23 reflect this reality.

"the essential emptiness"... I'm surprised, and perhaps a bit concerned, to
find that I have tapped into the existential despair of Anonymous Reviewer #2.

This study investigates how slip rates measured as *offset / time* of an offset
marker are influenced by earthquake cycle variation. It is *not* a paper
discussing methods or uncertainties invovlved in the calculation of the mean
and standard deviation of earthquake recurrence intervals or displacements
from paleoseismology. Indeed, if "There is an extensive statistical
literature..." (comment [B2b]) on this topic, why should I rehash it?

I fully agree that a typical paleoseismological study cannot resolve the mean
and standard deviation with a high degree of accuracy and I never stated
otherwise. Partly for this reason, by using CVs of 0.5, 1, and 2, I cover a
pretty wide range of values that can give some idea of the expected behavior of
the almost any fault system if the basic recurrence distributions can be
estimated or assumed.

A discussion of 'guesstimating' the $CV$ is actually already present in the
manuscript, starting on p.8, l. 31 in the original submission. However, I do
not give an in-depth treatment of estimating the $CV$ on a fault--this topic
would be suitable in a paper on interpreting field methods or a general
overview of statistical paleoseismology, perhaps, but not in a numerical
simulation paper that takes values of the $CV$ *a priori*.

@@

*Changes: None.*

> [B5a] It is not  obvious that the author has material experience words
> “aleatoric” and “epistemic”. Line 1, “aleatoric uncertainty” is a
> contradiction in terms.

It's really not a contradiciton in terms. 'Aleatory' does not mean 'the absence
of uncertainty', which it would have to in order to contradict the term
'uncertainty'.

In any case, the term 'aleatoric uncertainty' is commonly understood and used by
statisticians in addition to us proles.  See for example O'Hagan, T. (2004). Dicing with the unknown. *Significance*, 1(3), 132-133.

@@

*Changes: None.*

> [B5b] Bird, Zechar and Frankel all know better than to use the method the
> author alleges in lines 21 and 22 to arrive at epistemic uncertainty in slip
> rate.  They would more likely consider the allegation a misreading of their
> work.

This is a very perplexing comment. The line in question states, "The
uncertainty in the resulting slip estimate is typically treated as epistemic,
and quantified through the propagation of the measurement uncertainties on
the offset and time quantities (e.g., Bird, 2007; Zechar and Frankel, 2009)."
Both of the papers referenced are focused on estimating fault slip rates (or
*long-term offset rates* in Bird's terminology) that incorporate the measurement
uncertainty in both age and offset measurements. Both of their methods are based
in a convolution of the probability distributions for age and offset of the
marker features (i.e., the data with their uncertainties). Perhaps the reviewer
takes 'the propagation of the measurement uncertainties...' to mean that it is
done through adding them in quadrature to the final rate, but I clearly do not
specify the methods used here and the reviewer may want to reign in the
assumptions.

@@

Sorry!

*No changes.*

> [B5d] p2, L5:  A perturbation in slip rate would mean it was
> slipping at rate X, then changes to Y.

Yes. This is both the intended and the obvious interpretation of the sentence.
The slip rate at the fault trace varies because of variability in the frequency
and magnitude of accumulating offsets.

@@

*No changes.*

> [B5e] p.2, L14-17 have careful paleoseismologists doing reasonable things in
> one sentence, then imply they would make plainly rookie mistakes in the next.

The paragraph in question states that the open interval since the last
earthquake is often taken into consideration, however the variability in closed
intervals is not nearly as often taken into consideration--this is why I have
written the paper. Should I not state that paleoseismologists and
neotectonicists who consider the effects of the terminal open interval are
careful? I don't feel like being less generous. Should I not point out that
variability within the closed intervals is important and rarely considered?  No,
I think that I should. The first reference in this paragraph makes both the
'reasonable' decision and then the 'rookie mistake'. I am not simply making this
stuff up.

@@

Sorry!

*No changes.*
*Changes: None.*

> [B6a] p.4, L15-23: The descriptions of the lognormal variables here give one
> pause. First, log-normal parameters do not have units.

This comment is completely off the mark. The shape and scale parameters of a
lognormal distribution are indeed unitless, as they are the mean and standard
deviation of $\ln L$, where $L$ is a lognormal distribution (this transforms the
lognormal distribution into a normal distribution).

However, the mean and
standard deviation of the actual lognormal distribution $L$ are in the units of
the distribution itself. Nowhere in the manuscript did I state that $\mu$ and
$\sigma$, which are clearly defined as the mean and standard deviation of any
distribution in question, are the shape and scale parameters of $L$.  They are
simply the mean and standard deviation of the distribution.

@@

There is certainly a lot of autocorrelation of the samples if one were to
look at them sequentially. Nonetheless, even if sequential samples are
correlated, given enough samples, the final distribution of the samples won't
depend on the autocorrelation in the samples (this principle underlies the
family of powerful and commonly-used numerical methods based on Markov Chain
Monte Carlo sampling, for instance). This is why I use 2,000,000 years of
simulation: it ensures that the sampling is ample enough to not worry that
the serially-correlated samples have not sufficiently sampled the
distribution. I would have to use a much longer simulation (tens to hundreds
of millions of years) otherwise.

*Changes: None.*

> [B8a] p.6, L26:  Starts a narrative of the consequences of the standard error,
> as though the standard error was never invented.

I have added a brief discussion of the standard error, which is an OK
approximation of the results beyond 5-10 mean earthquake cycles (once the
systematic bias in very short time windows has gone away), but the asymmetry of
the results are not well approximated by the standard error. One could hack on
it a bit by log-transforming the data, applying the standard error of the
log-transformed distributions, and then anti-log transforming it, but that seems
a bit much.

@@

*Changes: None.*

> [B8b] The fuzzy, back-of-the-envelope estimates start to get thick here. Real
> uncertainty estimates would serve better.

The uncertainty estimates at a range of percentiles are available in Figure 5,
and the uncertainties for a slip rate measurement are in Figure 6 and Table 1.

@@

*Changes: None.*

> [B9a] p.7, L1-3: Two observations: First, as written, the practicing
> geologist is being asked to believe that 60 earthquake cycles have passed
> with zero displacement. I can guess what was intended, but should not have to.

Actually, the naive reading of the sentence, that 60 *mean* earthquake cycles
passed without an earthquake on a fault in the simulation, is exactly what is
intended.

@@

*No changes.*

### 2018-11-26 manuscript/reviews/response_letter.md

> [B9b] Second, what probability is associated with this 60-cycle thing?  I ask
> because practicing hazard geologists have to make estimates, and give weights
> to extreme events.  What is the probability of 60 cycles, a CV of 2.0, ...?
> Hard to imagine that the author  has thought much about what these results
> would mean or how to use them if they were true.

I happen to be a practicing hazard geologist who is responsible for the
implementation of fault sources for probabilistic seismic hazard analysis, so
I have actually considered this. It's very important when deciding whether to
include faults that have very clear bedrock geomorphic signatures (say, from
rapid deformation in the Miocene) and are favorably oriented for slip in the
current stress regime, but lack evidence of Quaternary deformation. Are they to
be included in the source models, or not?  I tend to add them, with low slip
rates and high uncertainties, for just this reason. The analysis in this paper
shows that very long recurrence intervals are possible on active faults.

There are two probabilities that are asked here:
1. *The probability of a fault with a CV of 2.0 going 60 mean earthquake cycles
   without accruing much displacement.*

This is answered very easily using the equation for the lognormal probability distribution with a $CV$ of 2. The cumulative distribution function for
this recurrence distribution at 60,000 years (60 mean earthquake cycles) is
0.9994..., which means that there is a probability of about $p = 0.0006$ that
any given recurrence interval will exceed 60,000 years, or 60 mean earthquake
cycles. But the probability $P$ of a recurrence interval this long being present
in a sequence of *n* earthquakes is $P=1 - (1-p)^n$, and for 2000 earthquakes
(which is the approximate number in these simulations) the probability of a
single recurrence interval exceeding 60 mean earthquake cycles is 0.1 or 10%.
It becomes more likely than not that such a recurrence interval will be
present in a sequence at about 12,500 earthquakes. Are you going to see this
in a trench? Almost certainly not even if you were to have the appropriate
sedimentation to preserve the signal. But that does not mean that it's an
impossible or even improbable event in a long time series.

*Changes*: I have added the following sentence to the paragraph under
discussion: "It is highly unlikely that any given recurrence
interval will be this long, but given thousands of earthquakes over millions
of years, the chance of such an event occurring at least once is far more
likely." I don't find it necessary to go through the probability calculations
in the manuscript as I did here.

2. *The probability of a fault having a CV of 2.0.*

The best way to calculate this probability is basically to look at the
frequency of such a CV occurring in some sample of faults. I don't have a big
collection of faults with paleoseismic data to work with, and I'm not going
to put one together to satisfy a reviewer, but I do calculate that the
schematic, representative model for southeastern Australian intraplate faults
by Clark et al., 2017 (cited in the paper) has a CV of 2. Unfortunately, Clark
et al. cannot precisely date all of the earthquakes they infer so no real CV can
be calculated from their data, but this is *their* best guess as to how the
faults behave, based on their evidence for clusters of events with recurrence
intervals of estimated mean 8,000 years separated by quiescent intervals of
0.5-2 million years.

Similarly, the USGS Quaternary Faults and Folds dataset has a number of faults
that are thought to have last ruptured in the 'Middle and Late Quaternary' (<1.6
Ma) but not in the 'Late Quaternary' (<130,000 years). Disregarding potential
misclassifications of these faults, it seems reasonable to assume that a few of
them may be experiencing very long recurrence intervals and may have a CV
approaching 2. Many of these faults are along strike of fault segments with
Holocene ruptures and there is no obvious geological difference between the
segments with recent rupture and those without--the Steens fault zone in the
lovely Steens rift of southeast Oregon is an example.

### 2018-11-26 manuscript/eq_cycle_slip_rates.tex
done with revisions for now

Please note that in the construction of the cumulative displacement
histories, all samples are independent. This means that the duration of
any recurrence interval does not depend on the duration of the previous
or subsequent interval (in other words, there is no autocorrelation in
these series); the same applies to the displacement samples.
Furthermore, the magnitude of displacement is independent of the
corresponding recurrence interval. It is currently unknown to what
degree autocorrelation exists in real earthquake time and displacement
series, or how much correlation is present between recurrence intervals
and subsequent displacements. Autocorrelation in recurrence interval sequences 
is essentially unstudied,
though on the basis of preliminary, unreviewed analysis
\citep{styron_survival_2017}, I suspect that it is as important as
\(CV\). With respect to the amount of slip per event and loading time, the 
framework of elastic rebound
theory in its most basic form should predict some correspondence between
inter-event (loading) duration and slip magnitude, and this is included
(implicitly or explicitly) in oscillator models incorporating complete
stress or strain release in each earthquake\\
\citep[e.g.,][]{matthews_brownian_2002,dicaprio_post-seismic_2008} or in
any model where coseismic friction drops to zero, as this is
functionally equivalent (because \(f^a = \tau^a_s / \tau^a_n\), where
\(f^a\), \(\tau^a_s\) and \(\tau^a_n\) are respectively friction at
rupture arrest, shear stress at rupture arrest, and effective normal
stress at rupture arrest, zero friction implies zero shear stress, or
complete stress drop). Given a reasonably constant loading rate,
complete shear stress or strain release implies some proportionality
between the loading time and displacement. Nonetheless, this
correspondence is not found in the more extensive paleoseismic datasets,
such as those by \citet{benedetti_earthquake_2013} (or the correlation
may be negative as found by \citet{weldon_wrightwood_2004}), but the
number of paleoseismic datasets of sufficient size and quantity to
identify these effects with statistical significance are few indeed.

@@

The uncertainty in the estimated slip rates due to earthquake cycle
variability is estimated by taking a function, \(\hat{R}\), that
calculates the mean slip rate within a time window \(t\), and sliding it
along the displacement series. \(\hat{R}\) is calculated simply as

\begin{equation}
  \hat{R}(D_0,D_1,t) = \frac{D_1 - D_0}{t}\; ,
  \label{eqn-rate-calc}
\end{equation}

where \(D_0\) is the cumulative displacement at the beginning of the
time window, \(D_1\) is the cumulative displacement at the end of the
time window, and \(t\) is the length of the time window. The \(\hat{}\)
symbol signifies an estimate rather than the true value \(R\). By
sliding \(\hat{R}\) over the displacement series, a set of many samples
of \(\hat{R}\) is generated, so that we may analyze the distribution.
The number of samples is \(n = N - t + 1\), where \(N\) is the length of
the total series (2,000,000 in this study).

A major goal of this study is to provide an answer to the question,
\emph{How long should slip rates be measured over in order to estimate a
meaningful rate?}\\
This question will be answered by looking at the distribution in
\(\hat{R}\) as a function of \(t\). Fifty values of \(t\) from 500 years
to 100,000 years, logarithmically spaced, are used. Note that given
\(\mu\) of 1000 years, this translates to 0.5-100 mean numbers of
earthquakes in the window.

The results of these calculations are shown in Figure
\ref{slip_rate_envelopes}. It is clear that the total variability in the
estimated slip rates is initially quite high when \(t\) is short
(\textless{}10,000 years or \textasciitilde{}10 earthquakes).
Particularly when \(t\) is \textless{}5,000 years, the maximum rates are
a factor of 3 or more greater than the true rate \(R\), but the median
rates are lower than \(R\)---this means it is more likely that fewer
earthquakes are captured in the time window than naively expected given
the mean recurrence. As the median is lower than \(R\), most
measurements over these short timescales will underestimate the mean
rate, although not necessarily by much.

With longer \(t\), between 10,000--20,000 years (or 10--20 earthquakes)
the variation in the slip rate estimates stabilize to within \(\pm\)
100\% of the mean (Figure \ref{slip_rate_envelopes}) for all
distributions, though this happens most quickly in the periodic
distribution, and most slowly in the clustered distribution. In fact,
the only exception here is that the lower bound of the clustered
distribution can stay at zero for more than 60 mean earthquake cycles.
For rate estimates longer than several tens of mean earthquake cycles,
the variation decreases very slowly but progressively with increasing
window length.

@@

Third, the median rate before convergence at 5 \(\mu\) is greater than
the true rate, meaning that in most cases, very short-term slip rate
measurements will underestimate the true slip rate; this is a systematic
bias. This is a particular concern for paleoseismological slip rate
estimates, where there are rarely more than 5 events in any given
trench. However, the median is less than 2 times \(\hat{R}\) following
just 2-3 events, so the systematic bias is unlikely to be much greater
than the measurement uncertainties on the age or offset of the events.

@@

However, it is unlikely that the values for the recurrence distributions
and the number of earthquakes that have transpired are sufficently known
to make the calculation, unless the fault has received sufficient
paleoseismic and neotectonic study. As formal hypothesis testing may not
be possible given typical slip rate datasets, an informal way of gauging
the likelihood of a slip rate change is to crudely estimate
(`guesstimate') the number of possible earthquakes and the recurrence
distribution, and then use the closest values in Table
\ref{ep_unc_table} with propagated measurement uncertainty to evaluate
the amount of overlap between the two slip rate estimates. If the
overlap between the distributions is a small fraction of the total range
of the distributions, then it is likely that a real slip rate change
occurred. This is clearly not appropriate for a real hypothesis test,
but it may aid researchers in developing ideas or intuition about the
behavior of a given fault.

>>>>

Please note that in the construction of the cumulative displacement
histories, all samples are independent. This means that the duration of any
recurrence interval does not depend on the duration of the previous or
subsequent interval (in other words, there is no autocorrelation in these
series); the same applies to the displacement samples. Furthermore, the
magnitude of displacement is independent of the corresponding recurrence
interval. It is currently unknown to what degree autocorrelation exists in
real earthquake time and displacement series, or how much correlation is
present between recurrence intervals and subsequent displacements.
Autocorrelation in recurrence interval sequences is essentially unstudied,
though on the basis of preliminary, unreviewed analysis
\citep{styron_survival_2017}, I suspect that it is as important as \(CV\).
With respect to the amount of slip per event and loading time, the framework
of elastic rebound theory in its most basic form should predict some
correspondence between inter-event (loading) duration and slip magnitude, and
this is included (implicitly or explicitly) in oscillator models
incorporating complete stress or strain release in each earthquake
\citep[e.g.,][]{matthews_brownian_2002,dicaprio_post-seismic_2008} or in any
model where coseismic friction drops to zero, as this is functionally
equivalent (because \(f^a = \tau^a_s / \tau^a_n\), where \(f^a\),
\(\tau^a_s\) and \(\tau^a_n\) are respectively friction at rupture arrest,
shear stress at rupture arrest, and effective normal stress at rupture
arrest, zero friction implies zero shear stress, or complete stress drop).
Given a reasonably constant loading rate, complete shear stress or strain
release implies some proportionality between the loading time and
displacement. Nonetheless, this correspondence is not found in the more
extensive paleoseismic datasets, such as those by
\citet{benedetti_earthquake_2013} (or the correlation may be negative as
found by \citet{weldon_wrightwood_2004}), but the number of paleoseismic
datasets of sufficient size and cquality to identify these effects with
statistical significance are few indeed.

@@

The uncertainty in the estimated slip rates due to earthquake cycle
variability is estimated by taking a function, \(\hat{R}\), that
calculates the mean slip rate within a time window \(t\), and sliding it
along the displacement series. \(\hat{R}\) is calculated simply as
\begin{equation}
  \hat{R}(D_0,D_1,t) = \frac{D_1 - D_0}{t}\; ,
  \label{eqn-rate-calc}
\end{equation}
where \(D_0\) is the cumulative displacement at the beginning of the
time window, \(D_1\) is the cumulative displacement at the end of the
time window, and \(t\) is the length of the time window. The \(\hat{}\)
symbol signifies an estimate rather than the true value \(R\). This slip rate
estimation method is intended to represent a neotectonic-style slip rate
estimate in which the number of earthquakes that have contributed to the
observed deformation is unknown, as are the durations of the open intervals that
bound the time window (one of which preceeds deposition of the marker unit, and
one is the time since the most recent earthquake and the measurement time).
By sliding \(\hat{R}\) over the displacement series, a set of many samples of
\(\hat{R}\) is generated, so that we may analyze the distribution. The number
of samples is \(n = N - t + 1\), where \(N\) is the length of the total
series (2,000,000 in this study).

A major goal of this study is to provide an answer to the question, \emph{How
long should slip rates be measured over in order to estimate a meaningful
rate?} This question will be answered by looking at the distribution in
\(\hat{R}\) as a function of \(t\). Fifty values of \(t\) from 500 years to
100,000 years, logarithmically spaced, are used. Note that given \(\mu\) of
1000 years, this translates to 0.5-100 mean numbers of earthquakes in the
window.

The results of these calculations are shown in Figure
\ref{slip_rate_envelopes} for up to 60 mean earthquake cycles. It is clear
that the total variability in the estimated slip rates is initially quite
high when \(t\) is short (\textless{}10,000 years or \textasciitilde{}10
earthquakes). Particularly when \(t\) is \textless{}5,000 years, the maximum
rates are a factor of 3 or more greater than the true rate \(R\), but the
median rates are lower than \(R\)---this means it is more likely that fewer
earthquakes are captured in the time window than naively expected given the
mean recurrence, and that the time contained in the open intervals is a
substantial fraction of the total time window. As the median is lower than
\(R\), most measurements over these short timescales will underestimate the
mean rate, although not necessarily by much.

With longer \(t\), between 10,000--20,000 years (or 10--20 earthquakes)
the variation in the slip rate estimates stabilize to within \(\pm\)
100\% of the mean (Figure \ref{slip_rate_envelopes}) for all
distributions, though this happens most quickly in the periodic
distribution, and most slowly in the clustered distribution. In fact,
the only exception here is that the lower bound of the clustered
distribution can stay at zero for more than 60 mean earthquake cycles. It is
highly unlikely that any given recurrence interval will be this long, but
given thousands of earthquakes over millions of years, the chance of such an
event occurring at least once is far more likely. For rate estimates longer
than several tens of mean earthquake cycles, the variation decreases very
slowly but progressively with increasing window length.

Note that with a measurement time exceeding 5-10 mean earthquake cycles, the
standard error ($n / \sqrt{(\sigma_D^2 + \sigma_r^2)}$, where $\sigma_D$ is
the standard deviation of displacement, and $\sigma_r$ is the standard
deviation of the recurrence interval) is a reasonable approximation for the
standard deviation of the variability in the slip rates due to
earthquake-cycle stochasticity, and shows broadly similar decreasing
variation with increasing earthquake cycles. However, the standard error is
symmetrical, though the variability displayed here is asymmetrical due to the
asymmetry of the recurrence interval and displacement distributions.

@@

Third, the median rate before convergence at 5 \(\mu\) is greater than
the true rate, meaning that in most cases, very short-term slip rate
measurements will underestimate the true slip rate; this is a systematic
bias. This is a particular concern for paleoseismological slip rate
estimates, where there are rarely more than 5 events in any given
trench, though this bias will be decreased if slip rate measurements are made
from closed intervals only. However, the median is less than 2 times
\(\hat{R}\) following just 2-3 events, so the systematic bias is unlikely to
be much greater than the measurement uncertainties on the age or offset of
the events.

@@

However, it is unlikely that the values for the recurrence distributions
and the number of earthquakes that have transpired are sufficently known
to make the calculation, unless the fault has received in-depth
paleoseismic and neotectonic study. As formal hypothesis testing may not
be possible given typical slip rate datasets, an informal way of gauging
the likelihood of a slip rate change is to crudely estimate
(`guesstimate') the number of possible earthquakes and the recurrence
distribution, and then use the closest values in Table
\ref{ep_unc_table} with propagated measurement uncertainty to evaluate
the amount of overlap between the two slip rate estimates. If the
overlap between the distributions is a small fraction of the total range
of the distributions, then it is likely that a real slip rate change
occurred. This is clearly not appropriate for a real hypothesis test,
but it may aid researchers in developing ideas or intuition about the
behavior of a given fault.
